A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-11-09 analysis/paper/paper.Rmd
Further changes from peer review and final comments from supervisors

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinterland and inland regions, where sites are believed to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf. @solheim2021]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the development of settled farmsteads is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2020; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a return to foraging and complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988; see e.g. @bergsvik2002; @bergsvik2012 for similarly nuanced considerations of the coastal settlement of Neolithic western Norway]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit (the highest elevation of the sea after the retreat of the ice) were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [see e.g. @berg-hansen2009; @eskeland2017; @nummedal1923; @simpson2009]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @berg-hansen2022, 644; @ilves2011].

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021].

As the displacement curves and their trajectory are quite complex constructs and the integrated result of both expert knowledge and more objectively quantifiable parameters, the geologists that have undertaken the studies have not found reason to assign variable uncertainty within the confidence envelopes of the displacement curves [@sørensen2014, 44; @romundset2018, 187]. The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

@@

Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level. Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.english(r2a0)` has simulation results for vertical distance that includes zero. The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m. While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE. Of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r r3a25` sites have all vertical distance results above 25m. However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline. The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.  @prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological developments suggested in the literature.

@@

This comparison is useful for illustrating both how the method has previously been employed, and for revealing nuances of the implementation that is proposed here. However, the comparison is also unfair to the previously proposed dates for a few reasons. First, the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Thirdly, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Additionally, the dates are often stated to be the result of also considering artefact typology and characteristics of local topography to inform the likely elevation of the sea when the site was in use---although precisely how these are weighted and used to inform the suggested date is often not as clear.

With a couple exceptions, the previously hypothesised dates and the ones achieved here appear to roughly correspond when it comes to the start date for the occupation of the sites. The most clear difference mainly pertains to the fact that the previously proposed date ranges are, without exception, more constrained than the 95% HDRs resulting from the proposed method. Considering the right skew of the probability mass functions underlying the 95% HDRs and the general overlap for the start dates, these results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success. This also follows from the fact that these dates have typically informed research in an approximate manner [although see e.g. @roalkvam2022].

With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision. While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method. The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates. In some cases the proposed method provides a fairly precise date and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf. Figure \@ref(fig:map-iso)B). However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology. While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed. This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.

@@

The most significant finding of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close proximity of sites and the shoreline until the transition to the Neolithic at c. 4000 BCE, after which a few sites are situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c. 2500 BCE. This development is in clear agreement with the literature. Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed. Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves accounting for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived gamma distribution with a shape of `r round(gammafit$estimate[1], 3)` and scale of `r round(gammafit$estimate[2], 3)` (Figure \@ref(fig:results2)A) to characterise this relationship. Furthermore, while this remains to be formalised and explored further, it was also showed how the method could potentially be improved by including more information on both the topographic location of the sites and other temporal data. To the degree that making such a distinction is useful, this could be derived from assessments of both a qualitative and quantitative nature, with Bayesian inference forming a natural framework for integrating such considerations [e.g. @buck1996; @otarola-castillo2023]. As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date, the impact of such additional information will also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted. For example, from Figure \@ref(fig:results2) it is clear that the Mesolithic sites have generally been located closer to the shoreline than the later sites. It was not attempted to explore this further here, given the constrained sample size and the good accuracy that was achieved with the parameters in use. However, the future addition of more data might give justification for using different models or parameter settings when dating sites from certain time intervals. Furthermore, following from its behavioural nature, it would also seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline the sites were located. This is illustrated here by the site phases associated with agricultural activity, marked in Figure \@ref(fig:results), which were all found to be located some distance from the sea. A wide range of different behavioural dimensions could potentially provide nuance to how the method should be applied. Other factors related to the topographic location of the sites could also be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @blankholm2020; @bondevik2019]. The potential of exploring dimensions such as these was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline. If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available.

@@

As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g. @nyland2020], the proposed implementation and parametrisation of shoreline dating can not be expected to be directly applicable elsewhere. When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g. @svendsen1987], this means that the accuracy and precision of the method can also be expected to vary. However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible. Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was here simply used to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.

>>>>

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b; @mansrud2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinterland and inland regions, where sites are believed to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf. @solheim2021]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the development of settled farmsteads is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2020; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a return to foraging and complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988; see e.g. @bergsvik2002; @bergsvik2012 for similarly nuanced considerations of the coastal settlement of Neolithic western Norway]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shoreline with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit (the highest elevation of the sea after the retreat of the ice) were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [see e.g. @berg-hansen2009; @eskeland2017; @nummedal1923; @simpson2009]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore-bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @berg-hansen2022, 644; @ilves2011].

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations below the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021].

As the displacement curves and their trajectory are quite complex constructs and the integrated result of both expert knowledge and more objectively quantifiable parameters, the geologists that have undertaken the studies have not found reason to assign variable uncertainty within the confidence envelopes of the displacement curves [@sørensen2014, 44; @romundset2018, 187]. The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well-defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located below sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

@@

Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level. Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.english(r2a0)` has simulation results for vertical distance that includes zero. The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m. Furthermore, some deviation is evident from just after 4000 BCE we well. The sample size is limited, but of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r as.english(r3a25)` sites have all vertical distance results above 25m. However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline. The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.  @prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological developments suggested in the literature.

@@

This comparison is useful for illustrating both how the method has previously been employed, and for revealing nuances of the implementation that is proposed here. However, the comparison is also unfair to the previously proposed dates for a few reasons. First, the dates provided in the reports are typically stated to be a very rough estimate and are sometimes given as a point estimate with an undefined, but implied or explicit uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Thirdly, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Additionally, the dates are often stated to be the result of also considering artefact typology and characteristics of local topography to inform the likely elevation of the sea when the site was in use---although precisely how these are weighted and used to inform the suggested date is often not as clear.

With a few exceptions, the previously hypothesised dates and the ones achieved here appear to roughly correspond when it comes to the start date for the occupation of the sites. The clearest difference mainly pertains to the fact that the previously proposed date ranges are, without exception, more constrained than the 95% HDRs resulting from the proposed method. Considering the right skew of the probability mass functions underlying the 95% HDRs and the general overlap for the start dates, these results could, with some danger of circularity, suggest that shoreline dating has generally been applied with a reasonable degree of success. This also follows from the fact that these dates have typically informed research in an approximate manner [although see e.g. @roalkvam2022].

With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision. While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method. The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates. In some cases the proposed method provides a relatively precise date and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf. Figure \@ref(fig:map-iso)B). However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology. While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed. This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.

@@

The most significant finding of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close proximity of sites and the shoreline until the transition to the Neolithic at c. 4000 BCE, after which a few sites are situated some distance from the sea, followed by a more decisive break at the transition to the Late Neolithic at c. 2500 BCE. This development is in clear agreement with the literature. Furthermore, based on the quantitative nature of these findings, an initial formulation of a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed. Apart from taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, this involves accounting for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived gamma distribution with a shape of `r round(gammafit$estimate[1], 3)` and scale of `r round(gammafit$estimate[2], 3)` (Figure \@ref(fig:results2)A) to characterise this relationship. Furthermore, while this remains to be formalised and explored further, it was also demonstrated how the method could potentially be improved by including more information on both the topographic location of the sites and other temporal data. To the degree that making such a distinction is useful, this could be derived from assessments of both a qualitative and quantitative nature, with Bayesian inference forming a natural framework for integrating such considerations [e.g. @buck1996; @otarola-castillo2023]. As the precision of the method is both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date, the impact of such additional information will also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted. For example, from Figure \@ref(fig:results2) it is clear that the Mesolithic sites have generally been located closer to the shoreline than the later sites. It was not attempted to explore this further here, given the constrained sample size and the accuracy that was achieved with the parameters in use. However, the future addition of more data might give justification for using different models or parameter settings when dating sites from certain time intervals. Furthermore, following from its behavioural nature, it is also likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline the sites were located. This is illustrated here by the site phases associated with agricultural activity, marked in Figure \@ref(fig:results), which were all found to be located some distance from the sea. A wide range of different behavioural dimensions could potentially provide nuance to how the method should be applied.

Other factors related to the topographic location of the sites could also be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range and potential impact of storm surges could also have implications for the location of a site relative to the shoreline, depending on the topography [@helskog1978; @bondevik2019]. The potential of exploring such dimensions was also hinted at here with the estimation and cursory treatment of the horizontal and topographic distance to the shoreline. If patterns related to such locational patterns can be discerned and unpacked, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information beyond their location is available. A mention should also be made here of the fact that catastrophic events such as tsunamis might also be of relevance [e.g. @blankholm2020; @nielsen2020; @nyland2021]. Evidence for the impact of tsunamis in the Stone Age has not been identified in south-eastern Norway as of yet [see @romundset2015, 398], and might therefore not be of direct relevance to the coastal settlement in the region. However, the outburst flood resulting from the catastrophic drainage of the glacial lake Nedre Glomsjø around 8500--8000 BCE [@høgaas2019], located in Mid-Norway some 230 km north of present-day Oslo, could have had consequences for how the coast was utilised [@solheim2020b, 9].

@@

As the nature of the relationship between sites and sea is likely to vary temporally and geographically [e.g. @nyland2020], the proposed implementation and parametrisation of shoreline dating can not be expected to be directly applicable elsewhere. When this is combined with the fact that the rate of RSL-change also varies geographically and temporally [e.g. @svendsen1987], this means that the accuracy and precision of the method will also vary. However, the methodological framework used to evaluate the relationship between sites and sea is readily extendible to other regions of northern Scandinavia where reliable data on shoreline displacement is available, thus making such extensions feasible. Furthermore, the simulation approach used to integrate multiple sources of spatio-temporal uncertainty was used here to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.
